Kozen Categories

### 2025-11-11 kleenecat/kleenecat.tex
edits on distributive law section

\documentclass[acmsmall,anonymous,review,screen]{acmart}
\usepackage{mathpartir}
\usepackage{quiver}
\usepackage{enumitem}
\usepackage{wrapfig}
\usepackage{fancyvrb}
\usepackage{amsmath}
\usepackage{stmaryrd}

@@

\begin{document}
\title{Kleene Categories}
\author{Steven Schaefer}
\affiliation{
  \department{Electrical Engineering and Computer Science}
  \institution{University of Michigan}
  \country{USA}
}
\email{stschaef@umich.edu}

\author{Max S. New}
\affiliation{
  \department{Electrical Engineering and Computer Science}
  \institution{University of Michigan}
  \country{USA}
}
\email{maxsnew@umich.edu}

\author{Pedro H. Azevedo de Amorim}
\affiliation{
  \department{Department of Computer Science}
  \institution{University of Oxford}
  \country{UK}
}
\email{pedro.azevedo.de.amorim@cs.ox.ac.uk}

@@

\section{Kleene Algebra and Kleene Categories}

@@

Our goal is to provide a generalization of Kleene algebra that is
``proof-relevant'', and so can provide a foundation for parsing
regular expressions in the same way that Kleene algebra does for
recognition. To do this we need to move beyond the view of a Kleene
algebra as algebraic structure over sets, and instead as a kind of
structured \emph{poset}. We then generalize this to a Kleene
\emph{category}, where the morphisms of the category correspond to not
just language inclusion but \emph{formal reductions} of grammars, and
isomorphism is not language equivalence but instead \emph{strong
equivalence} of grammars.

@@

\begin{definition}
  A Kleene category is a distributive monoidal category $\cat{K}$ such
  that for every object $A$, there is a cons-list object
  $(A^*,\textrm{nil},\textrm{cons})$ such that $(A^*, \textrm{nil},
  \textrm{snoc})$ is a snoc-list object.
\end{definition}

As a sanity check, note that Kleene algebras are indeed examples of
Kleene categories.

\begin{lemma}
  A Kleene algebra is equivalent to a posetal Kleene category.
\end{lemma}

% This isn't an example of a Kleene category, it's an example of a model of conjunctive ordered logic
An unexpected example comes from the theory of substructural logics.

\begin{example}
  The opposite category of every Kleene category is a model of a variant of
  conjunctive ordered logic, where the Kleene star plays the role of the ``of
  course'' modality from substructural logics.
\end{example}

This proposed definition is sensible. After all, the proposed axioms
are a direct translation of the Kleene algebra axioms to a categorical
setting. Its most convoluted aspect is the axiomatization of the
Kleene star as a parameterized list and snoc object. If the Kleene
category $\cat{K}$ has more structure, then these isomorphisms hold
``for free''.

\begin{definition}
  A pre-Kleene category is a distributive monoidal category with a (non-parameterized) cons-list object.
\end{definition}

\begin{theorem}
  If $\cat K$ is a pre-Kleene category that is also biclosed, then it is a Kleene category.
\end{theorem}
\begin{proof}
  TODO: closed implies parameterized for free and that cons-list and snoc-list objects are equivalent.
\end{proof}

%% \begin{theorem}
%%   \label{th:kleeneclosed}
%%   Let $\cat{K}$ be a Kleene category such that it is also monoidal
%%   closed.  Then, the initial algebras isomorphisms hold automatically.
%% \end{theorem}
%% \begin{proof}
%%   We prove this by the unicity (up-to isomorphism) of initial
%%   algebras. Let $[hd, tl]: 1 + (\mu X.\, F_{A, 1}(X)) \otimes A \to
%%   (\mu X.\, F_{A, 1}(X))$ be the initial algebra structure of $(\mu
%%   X.\, F_{A, 1}(X))$ and consider the map $[hd, tl] : B + B \otimes
%%   (\mu X.\, F_{A, 1}(X)) \otimes A \to B\otimes (\mu X.\, F_{A,
%%     1}(X))$.

@@

%% \begin{theorem}
%%   Let $\cat{K}$ be a Kleene category and $A$ a discrete category.
%%   The functor category $[A, \cat{K}]$.
%%   (HOW GENERAL SHOULD THIS THEOREM BE? BY ASSUMING ENOUGH STRUCTURE,
%%   E.G. K = Set, THIS THEOREM BECOMES SIMPLE TO PROVE)
%% \end{theorem}
\begin{theorem}
  If $\cat{C}$ is a locally small monoidal category, then
  $\Set^{\cat{C}}$ is a Kleene category.
\end{theorem}
\begin{proof}
  By the lemma above, $\Set^{\cat{C}}$ is monoidal closed, and since it
  is a presheaf category, it has coproducts, and since the tensor
  is a left adjoint, it preserves colimits and, therefore, it is
  a distributive category.

@@

\begin{corollary}
  For every alphabet $\Sigma$, the presheaf category $\grammar$
  is a Kleene category.
\end{corollary}
\begin{proof}
  Note that string concatenation and the empty string make the
  discrete category $\Sigma^*$ a strict monoidal category.
\end{proof}

\begin{definition}
  Let $K$ be a Kleene category over alphabet $\Sigma$. This
  induces a strong monoidal functor $\Sigma^* \to K$. The
  nerve functor $w \mapsto K(w, -)$ is the grammar interpretation
  functor of $K$.
\end{definition}

@@

Assuming that the underlying category $\cat{C}$ is not discrete,
arrows $f : X \to Y$ in $\cat{C}$ should be thought of as a proof that
the state $X$ is related to the $Y$. This can be seen as a
generalization of simulation relations. After all, functoriality of $F
: \cat{C} \times \Sigma \to \cat{C}$ implies that $F(f, id_c): F(X, c)
\to F(Y, c)$, i.e. after one transition, the end states will also be
related.

@@

\section{Proposed construction for free Kleene categories}

@@

Let $\cat{C}$ be a category. It is more or less straightforward to
define the free Kleene category $\FKC(\cat{C})$. We can do that by
using the internal language of Kleene categories. Concretely, it
will be a type theory with (non-commutative) tensors, sum types and
a primitive type of lists $L(A)$ for every $A$. By type theoretic reasons, this
structure forces the coproduct to distribute over the tensor, where we
can write the morphism $A \otimes (B + C) \to (A \otimes B) + (A
\otimes C) $ as the term
\begin{align*}
  &z \vdash \mathsf{let}\, x \otimes y = z\, \mathsf{in}\\
  &\mathsf{case}\, y \, \mathsf{with}\\
  &| \mathsf{in}_1\, y_1 \Rightarrow \mathsf{in}_1\, (x \otimes y_1)\\
  &| \mathsf{in}_2\,y_2\Rightarrow \mathsf{in}_2 \,(x \otimes y_2)\\
  &\mathsf{end}
\end{align*}
%
The list type includes primitive terms for $\textrm{cons}$,
$\textrm{nil}$, $\textrm{foldr}$ and $\textrm{foldl}$ where
$\textrm{foldr}$ and $\textrm{foldl}$ allow for a context to be used
in the construction of the nil case, but no free variables can be used
in the cons case:
\begin{mathpar}
  \inferrule
  {\Gamma \vdash L : A^* \and \Delta \vdash N : B \and a:A,b:B \vdash M : B
  }
  {\Gamma,\Delta \vdash \textrm{foldr}(L,N,a.b.M) : B}
\end{mathpar}

@@

The proof that this is a Kleene algebra is boring, but unsurprising.
Because the axiomatization of Kleene algebras over an alphabet $\Sigma$
is a Horn theory, the category of Kleene algebras and Kleene algebra
morphisms is monadic over $\Set$. Let $E_\Sigma : \Set \to \Set$ be the 
Kleene algebra monad.

@@

We now want to prove that the distributive law above generalizes to
the Kleene category case. First, we have to define a natural
transformation $K(\Set \times (-)^\Sigma) \to \Set \times (K -
)^\Sigma$ making the appropriate diagrams commute. We decompose the
natural transformation into two components.  The component $K(\Set
\times (-)^\Sigma) \to \Set$ is by equipping $\Set$ with a Kleene
category structure and then using the universal property of the free
Kleene category. The Kleene category structure on $\Set$ is the usual
(co)cartesian structure and $X^*$ are the lists on $X$. Therefore, it
suffices to define a natural transformation $\Set \times (-)^\Sigma \to \Set$,
which we define as the first projection.

For the component $K(\Set \times (-)^\Sigma) \to (K - )^\Sigma$,
we will once again use the universal property of free Kleene
categories. We begin by showing that the ``non-standard''
structure defined in the previous section can be extended to the
proof-relevant setting.

\pa{This is where a better understanding of what this ``non-standard''
  construction is actually doing could come in handy. Showing that the
  monoidal product satisfies all of the coherence diagrams is really gnarly.}
\begin{lemma}
  Let $\cat{K}$ and $\cat{L}$ be two Kleene categories, $F : \cat{L} \to \cat{K}$
  and $G : \cat{L} \to \cat{K}$ two Kleene functors.
  We can equip $\cat{K} \times \cat{L}$ with the following Kleene
  category structure:
  \begin{itemize}
  \item $(A, X) + (B, Y) = (A + B, X + Y)$
  \item $0 = (0, 0)$
  \item $(A, X) \otimes (B, Y) = (A \otimes F(Y) + G(X) \otimes B, X \otimes Y)$
  \item $I = (0, I)$
  \item $(A, X)^* = (A \otimes F(X^*), X^*)$
  \end{itemize}
\end{lemma}

Therefore we equip $K(\cat{C})^\Sigma \times K(\cat{C})$ with the
Kleene category structure given by the lemma above.
%% \begin{proof}
%%   Since the morphism structure is defined componentwise, it is
%%   immediate to show that the proposed cocartesian structure is
%%   valid. For the monoidal structure, the bifunctoriality of $\otimes$
%%   follows from the bifunctoriality of the monoidal structures
%%   $\otimes$ and $+$ in $\cat{K}$. Next, we define the coherence
%%   isomorphisms. As suggested by the action on the $\cat{L}$ component,
%%   such isomorphisms are the same as in $\cat{L}$. The $\cat{K}$
%%   components of these morphisms are more interesting:
%% \end{proof}

@@

\section{Canonicity for Free Kleene Categories}

Let $\Sigma$ be a set. Then the ``standard semantics'' of the free
Kleene category $\FKC(\Sigma)$ is the unique functor $\sem{\cdot} :
\FKC(\Sigma) \to \Set^{\Sigma^*}$ preserving the (Day convolution)
Kleene category structure.

We first address whether we can generalize Kozen's completeness
theorems for Kleene algebras. Firstly, the fact that the Free Kleene
category is complete for \emph{language} inclusion and \emph{weak}
equivalence follows immediately from the fact that the free Kleene
algebra is equivalent to the posetal reflection of the free Kleene
category and the completeness theorem for Kleene algebras. That is,
for any objects $A, B \in \FKC(\Sigma)$, if $|\sem{A}| \subseteq
|\sem{B}|$ then there is a morphism $A \to B$ in $\FKC(\Sigma)$.

The immediate generalization to strong equivalence is false. For
example, we might ask that $\sem{\cdot}$ is a \emph{full} functor,
that any semantic reduction of the formal grammars is the denotation
of a morphism in the free Kleene category. But consider that the
regular grammar $1^*$ denotes a discrete grammar of natural numbers
and so morphisms $1^* \otimes \cdots \to 1^*$ are mapped into a set of
multi-argument functions on natural numbers. So completeness would
imply that all functions on natural numbers can be defined in the
syntax of the free Kleene category, which fails as all such functions
should be primitive recursive\footnote{flesh out the
details?}.

Failing completeness for definability of reductions, we could ask that
the \emph{equational theory} of free Kleene categories is
complete. This means that $\sem{\cdot}$ is a \emph{faithful} functor,
i.e., if two syntactic morphisms denote the same reduction $\sem{M} =
\sem{N}$, then they are provably equal $M = N$. I don't know if this
is true yet, but it seems unlikely to me.

@@

To do this we will explicitly construct an inverse natural
transformation $q : \sem{\cdot} \to N$ which we call ``quote'' because it quotes
a semantic parse tree back into the syntax. $q$ can be defined
explicitly by recursion on syntactic types and naturality proven by
induction over syntactic terms but we will develop a more structured
proof using the universal property of the free Kleene category.  To do
this we need to develop a Kleene category for which constructing a
functor from $\FKC(\Sigma)$ into it is equivalent to defining such a
natural transformation. We can do this by using the following
properties of \emph{comma categories}.

@@

Then we can prove the following
\begin{theorem}[Gluing for Kleene categories]
  Let $\cat K, \cat L, \cat M$ be Kleene categories and $\Phi : \cat K
  \to \cat M$ be a Kleene functor and $G : \cat L \to \cat M$ be a lax
  monoidal functor. Then $\Phi \downarrow G$ is a Kleene category, and
  the projections are (strict) Kleene functors.
\end{theorem}
\begin{proof}
  The monoidal and coproduct aspects of Kleene categories are handled
  by the previous lemma. We need to show that the monoidal structure
  is distributive and construct the Kleene star.

@@

Then $\sem {\cdot} : \FKC(\Sigma) \to \Set^{\Sigma^*}$ is a strict
Kleene functor and $N : \FKC(\Sigma) \to \Set^{\Sigma^*}$ is a lax
monoidal functor, and so $\sem{\cdot}\downarrow N$ is a Kleene
category and the projection to $\FKC(\Sigma)^2$ is a strict Kleene
functor. Then by the universal property of $\FKC(\Sigma)$, there is a
unique strict Kleene functor $q$ from $\FKC(\Sigma)$ to
$\sem{\cdot}\downarrow N$ and further the composition with the
projection to $\FKC(\Sigma)^2$ is strict Kleene and so $q$ is
equivalent to a natural transformation $\sem{\cdot} \Rightarrow N$.

@@

\begin{lemma}
  $u \circ q$ is the identity
\end{lemma}
\begin{proof}
  By the universal property of $\FKC(\Sigma)$, it suffices to show
  that $u \circ q$ and $\id$ correspond to (strict?) Kleene functors
  from $\FKC(\Sigma)$ to $\sem{\cdot} \downarrow \sem{\cdot}$. The
  identity part should be a general theorem, and verifying the Kleene
  category properties is done by cases.
\end{proof}

>>>>

\documentclass[sigplan,anonymous,review,screen]{acmart}
\usepackage{mathpartir}
\usepackage{quiver}
\usepackage{enumitem}
\usepackage{wrapfig}
\usepackage{fancyvrb}
\usepackage{amsmath}
\usepackage{stmaryrd}

@@

\begin{document}
\title{Kozen Categories}

@@

\section{Kleene Algebra and Kozen Categories}

@@

Our goal is to provide a generalization of Kleene algebra that is
``proof-relevant'', and so can provide a foundation for parsing
regular expressions in the same way that Kleene algebra does for
recognition. To do this we need to move beyond the view of a Kleene
algebra as algebraic structure over sets, and instead as a kind of
structured \emph{poset}. We then generalize this to a Kozen
\emph{category}, where the morphisms of the category correspond to not
just language inclusion but \emph{formal reductions} of grammars, and
isomorphism is not language equivalence but instead \emph{strong
equivalence} of grammars.

@@

\begin{definition}
  A Kozen category is a distributive monoidal category $\cat{K}$ such
  that for every object $A$, there is a cons-list object
  $(A^*,\textrm{nil},\textrm{cons})$ such that $(A^*, \textrm{nil},
  \textrm{snoc})$ is a snoc-list object.
\end{definition}

As a sanity check, note that Kleene algebras are indeed examples of
Kozen categories.

\begin{lemma}
  A Kleene algebra is equivalent to a posetal Kozen category.
\end{lemma}

% This isn't an example of a Kozen category, it's an example of a model of conjunctive ordered logic
An unexpected example comes from the theory of substructural logics.

\begin{example}
  The opposite category of every Kozen category is a model of a variant of
  conjunctive ordered logic, where the Kleene star plays the role of the ``of
  course'' modality from substructural logics.
\end{example}

This proposed definition is sensible. After all, the proposed axioms
are a direct translation of the Kleene algebra axioms to a categorical
setting. Its most convoluted aspect is the axiomatization of the
Kleene star as a parameterized list and snoc object. If the Kozen
category $\cat{K}$ has more structure, then these isomorphisms hold
``for free''.

\begin{definition}
  A pre-Kozen category is a distributive monoidal category with a (non-parameterized) cons-list object.
\end{definition}

\begin{theorem}
  If $\cat K$ is a pre-Kozen category that is also biclosed, then it is a Kozen category.
\end{theorem}
\begin{proof}
  TODO: closed implies parameterized for free and that cons-list and snoc-list objects are equivalent.
\end{proof}

%% \begin{theorem}
%%   \label{th:kleeneclosed}
%%   Let $\cat{K}$ be a Kozen category such that it is also monoidal
%%   closed.  Then, the initial algebras isomorphisms hold automatically.
%% \end{theorem}
%% \begin{proof}
%%   We prove this by the unicity (up-to isomorphism) of initial
%%   algebras. Let $[hd, tl]: 1 + (\mu X.\, F_{A, 1}(X)) \otimes A \to
%%   (\mu X.\, F_{A, 1}(X))$ be the initial algebra structure of $(\mu
%%   X.\, F_{A, 1}(X))$ and consider the map $[hd, tl] : B + B \otimes
%%   (\mu X.\, F_{A, 1}(X)) \otimes A \to B\otimes (\mu X.\, F_{A,
%%     1}(X))$.

@@

%% \begin{theorem}
%%   Let $\cat{K}$ be a Kleene category and $A$ a discrete category.
%%   The functor category $[A, \cat{K}]$.
%%   (HOW GENERAL SHOULD THIS THEOREM BE? BY ASSUMING ENOUGH STRUCTURE,
%%   E.G. K = Set, THIS THEOREM BECOMES SIMPLE TO PROVE)
%% \end{theorem}
\begin{theorem}
  If $\cat{C}$ is a locally small monoidal category, then
  $\Set^{\cat{C}}$ is a Kozen category.
\end{theorem}
\begin{proof}
  By the lemma above, $\Set^{\cat{C}}$ is monoidal closed, and since it
  is a presheaf category, it has coproducts, and since the tensor
  is a left adjoint, it preserves colimits and, therefore, it is
  a distributive category.

@@

\begin{corollary}
  For every alphabet $\Sigma$, the presheaf category $\grammar$
  is a Kozen category.
\end{corollary}
\begin{proof}
  Note that string concatenation and the empty string make the
  discrete category $\Sigma^*$ a strict monoidal category.
\end{proof}

\begin{definition}
  Let $K$ be a Kozen category over alphabet $\Sigma$. This
  induces a strong monoidal functor $\Sigma^* \to K$. The
  nerve functor $w \mapsto K(w, -)$ is the grammar interpretation
  functor of $K$.
\end{definition}

@@

Assuming that the underlying category $\cat{C}$ is not discrete,
arrows $f : X \to Y$ in $\cat{C}$ should be thought of as a proof that
the state $X$ is related to $Y$. This can be seen as a
generalization of simulation relations. After all, functoriality of $F
: \cat{C} \times \Sigma \to \cat{C}$ implies that $F(f, id_c): F(X, c)
\to F(Y, c)$, i.e. after one transition, the end states will also be
related. We now formalize this intuition

\begin{example}[Automata with simulation relation]
  Various notions of simulation for automata can be formulated using
  the language of fibered category theory and liftings of functors.
  Assume that we have a determinisitc automaton over a state space $S$
  with a simulation relation $\mathcal{R}$. We can see $S$ and $\mathcal{R}$
  as a thin category and equip it with a canonical $F$-coalgebra structure.
  \begin{itemize}
  \item The component $(S, \mathcal{R}) \to \Set$ maps an element of $S$ to the
    singleton set $\{*\}$ if it is an accepting state and to the empty set
    otherwise. The action on morphisms is defined using the simulation
    acceptance condition, i.e. if $(x, y) \in \mathcal{R}$, then
    $\alpha(x) \leq \alpha(y)$. Functoriality follows from initiality/terminality
    of the empty and singleton sets, respectively.
  \item The component $(S, \mathcal{R}) \times \Sigma \to (S, \mathcal{R})$
    is, as expected, the transition function $\delta : S \times \Sigma \to S$.
    The action on morphisms is given by the transition condition of simulation
    relations, i.e. if $(x, y) \in \mathcal{R})$ then for every $c \in \Sigma$
    $(\delta(x, c), \delta(y, c)) \in \mathcal{R}$.
  \end{itemize}
\end{example}

@@

\section{Proposed construction for free Kozen categories}

@@

Let $\cat{C}$ be a category. It is more or less straightforward to
define the free Kozen category $\FKC(\cat{C})$. We can do that by
using the internal language of Kozen categories. Concretely, it
will be a type theory with (non-commutative) tensors, sum types and
a primitive type of lists $L(A)$ for every $A$. By type theoretic reasons, this
structure forces the coproduct to distribute over the tensor, where we
can write the morphism $A \otimes (B + C) \to (A \otimes B) + (A
\otimes C) $ as the term
\begin{align*}
  &z \vdash \mathsf{let}\, x \otimes y = z\, \mathsf{in}\\
  &\mathsf{case}\, y \, \mathsf{with}\\
  &| \mathsf{in}_1\, y_1 \Rightarrow \mathsf{in}_1\, (x \otimes y_1)\\
  &| \mathsf{in}_2\,y_2\Rightarrow \mathsf{in}_2 \,(x \otimes y_2)\\
  &\mathsf{end}
\end{align*}
%
The list type includes primitive terms for $\textrm{cons}$,
$\textrm{nil}$, $\textrm{foldr}$ and $\textrm{foldl}$ where
$\textrm{foldr}$ and $\textrm{foldl}$ allow for a context to be used
in the construction of the nil case, but no free variables can be used
in the cons case:
\begin{mathpar}
  \inferrule
  {\Gamma \vdash L : A^* \and \Delta \vdash N : B \and a:A,b:B \vdash M : B
  }
  {\Gamma,\Delta \vdash \textrm{foldr}(L,N,a.b.M) : B}
\end{mathpar}

@@

The proof that this is a Kleene algebra is boring and unsurprising.
Because the axiomatization of Kleene algebras over an alphabet $\Sigma$
is a Horn theory, the category of Kleene algebras and Kleene algebra
morphisms is monadic over $\Set$. Let $E_\Sigma : \Set \to \Set$ be the 
Kleene algebra monad.

@@

We now want to prove that the distributive law above generalizes to
the Kozen category case. First, we have to define a natural
transformation $K(\Set \times (-)^\Sigma) \to \Set \times (K -
)^\Sigma$ making the appropriate diagrams commute. We decompose the
natural transformation into two components.  The component $K(\Set
\times (-)^\Sigma) \to \Set$ is by equipping $\Set$ with a Kozen
category structure and then using the universal property of the free
Kozen category. The Kozen category structure on $\Set$ is the usual
(co)cartesian structure and $X^*$ are the lists on $X$. Therefore, it
suffices to define a natural transformation $\Set \times (-)^\Sigma \to \Set$,
which we define as the first projection.

For the component $K(\Set \times (-)^\Sigma) \to (K - )^\Sigma$,
we will once again use the universal property of free Kozen
categories. We begin by showing that the ``non-standard''
structure defined in the previous section can be extended to the
proof-relevant setting.

\pa{This is where a better understanding of what this ``non-standard''
  construction is actually doing could come in handy. Showing that the
  monoidal product satisfies all of the coherence diagrams is really gnarly.}
\begin{lemma}
  Let $\cat{K}$ and $\cat{L}$ be two Kozen categories, $F : \cat{L} \to \cat{K}$
  and $G : \cat{L} \to \cat{K}$ two Kozen functors.
  We can equip $\cat{K} \times \cat{L}$ with the following Kozen
  category structure:
  \begin{itemize}
  \item $(A, X) + (B, Y) = (A + B, X + Y)$
  \item $0 = (0, 0)$
  \item $(A, X) \otimes (B, Y) = (A \otimes F(Y) + G(X) \otimes B, X \otimes Y)$
  \item $I = (0, I)$
  \item $(A, X)^* = (A \otimes F(X^*), X^*)$
  \end{itemize}
\end{lemma}

Therefore we equip $K(\cat{C})^\Sigma \times K(\cat{C})$ with the
Kozen category structure given by the lemma above.

  In our case, we want to instantiate $F$ with the unit for the reader
  monad $(-)^\Sigma$. The generalization of $G$ is a bit subtle, as it
  is not enough to simply output weather a state accepts the empty
  string or not, we have to carry around the the witnesses of such
  parse trees. Usin the fact that $2$ is the free Kleene algebra with
  zero generators on the empty alphabet, we instantiate $G$ to be a
  ``prunning'' operation that factors through the free Kozen category
  on the empty category. More concretely, we define a functor
  $K(\cat{C}) \to K(\cat{C})^\Sigma$ by first defining a functor
  $K(\cat{C}) \to K(\emptyset)$ by using the universal property of free
  Kozen categories and mapping $\cat{C}$ to the constant $0$ object of
  $K(\emptyset)$. The functor $K(\emptyset) \to K(\cat{C})^\Sigma$ is defined
  using the fact that left adjoints preserve colimits, so $K(\emptyset)$
  is initial in the category of Kozen categories.

%% \begin{proof}
%%   Since the morphism structure is defined componentwise, it is
%%   immediate to show that the proposed cocartesian structure is
%%   valid. For the monoidal structure, the bifunctoriality of $\otimes$
%%   follows from the bifunctoriality of the monoidal structures
%%   $\otimes$ and $+$ in $\cat{K}$. Next, we define the coherence
%%   isomorphisms. As suggested by the action on the $\cat{L}$ component,
%%   such isomorphisms are the same as in $\cat{L}$. The $\cat{K}$
%%   components of these morphisms are more interesting:
%% \end{proof}

@@

\section{Canonicity for Free Kozen Categories}

Let $\Sigma$ be a set. Then the ``standard semantics'' of the free
Kozen category $\FKC(\Sigma)$ is the unique functor $\sem{\cdot} :
\FKC(\Sigma) \to \Set^{\Sigma^*}$ preserving the (Day convolution)
Kozen category structure.

We first address whether we can generalize Kozen's completeness
theorems for Kleene algebras. Firstly, the fact that the Free Kozen
category is complete for \emph{language} inclusion and \emph{weak}
equivalence follows immediately from the fact that the free Kleene
algebra is equivalent to the posetal reflection of the free Kozen
category and the completeness theorem for Kleene algebras. That is,
for any objects $A, B \in \FKC(\Sigma)$, if $|\sem{A}| \subseteq
|\sem{B}|$ then there is a morphism $A \to B$ in $\FKC(\Sigma)$.

The immediate generalization to strong equivalence is false. For
example, we might ask that $\sem{\cdot}$ is a \emph{full} functor,
that any semantic reduction of the formal grammars is the denotation
of a morphism in the free Kozen category. But consider that the
regular grammar $1^*$ denotes a discrete grammar of natural numbers
and so morphisms $1^* \otimes \cdots \to 1^*$ are mapped into a set of
multi-argument functions on natural numbers. So completeness would
imply that all functions on natural numbers can be defined in the
syntax of the free Kozen category, which fails as all such functions
should be primitive recursive\footnote{flesh out the
details?}.

Failing completeness for definability of reductions, we could ask that
the \emph{equational theory} of free Kozen categories is
complete. This means that $\sem{\cdot}$ is a \emph{faithful} functor,
i.e., if two syntactic morphisms denote the same reduction $\sem{M} =
\sem{N}$, then they are provably equal $M = N$. I don't know if this
is true yet, but it seems unlikely to me.

@@

To do this we will explicitly construct an inverse natural
transformation $q : \sem{\cdot} \to N$ which we call ``quote'' because it quotes
a semantic parse tree back into the syntax. $q$ can be defined
explicitly by recursion on syntactic types and naturality proven by
induction over syntactic terms but we will develop a more structured
proof using the universal property of the free Kozen category.  To do
this we need to develop a Kozen category for which constructing a
functor from $\FKC(\Sigma)$ into it is equivalent to defining such a
natural transformation. We can do this by using the following
properties of \emph{comma categories}.

@@

Then we can prove the following
\begin{theorem}[Gluing for Kozen categories]
  Let $\cat K, \cat L, \cat M$ be Kozen categories and $\Phi : \cat K
  \to \cat M$ be a Kozen functor and $G : \cat L \to \cat M$ be a lax
  monoidal functor. Then $\Phi \downarrow G$ is a Kozen category, and
  the projections are (strict) Kozen functors.
\end{theorem}
\begin{proof}
  The monoidal and coproduct aspects of Kozen categories are handled
  by the previous lemma. We need to show that the monoidal structure
  is distributive and construct the Kleene star.

@@

Then $\sem {\cdot} : \FKC(\Sigma) \to \Set^{\Sigma^*}$ is a strict
Kozen functor and $N : \FKC(\Sigma) \to \Set^{\Sigma^*}$ is a lax
monoidal functor, and so $\sem{\cdot}\downarrow N$ is a Kozen
category and the projection to $\FKC(\Sigma)^2$ is a strict Kozen
functor. Then by the universal property of $\FKC(\Sigma)$, there is a
unique strict Kozen functor $q$ from $\FKC(\Sigma)$ to
$\sem{\cdot}\downarrow N$ and further the composition with the
projection to $\FKC(\Sigma)^2$ is strict Kozen and so $q$ is
equivalent to a natural transformation $\sem{\cdot} \Rightarrow N$.

@@

\begin{lemma}
  $u \circ q$ is the identity
\end{lemma}
\begin{proof}
  By the universal property of $\FKC(\Sigma)$, it suffices to show
  that $u \circ q$ and $\id$ correspond to (strict?) Kozen functors
  from $\FKC(\Sigma)$ to $\sem{\cdot} \downarrow \sem{\cdot}$. The
  identity part should be a general theorem, and verifying the Kozen
  category properties is done by cases.
\end{proof}
